Optimal updating factor in adaptive flat-distribution-sampling simulations

### 2017-04-26 doc/invt.tex
update invt.c

\begin{abstract}
  We present a method of computing the optimal schedule
  of the updating magnitude
  for a class of free energy calculation methods including
  the Wang-Landau (WL) algorithm and metadynamics.
  %
  These methods rely on adaptive construction of
  a bias potential that offsets
  the potential of mean force by histogram-based adaptive updates.
  %
  The updating magnitude should decrease over time
  to reduce the noise in the bias potential,
  and one may speed up the convergence by choosing an optimal schedule
  of the updating magnitude.
  %
  %Here, we will show that the optimal schedule
  %can be derived
  %from a mechanical analogy, in which
  %the schedule corresponds to
  %the velocity of a free particle with a position-dependent mass
  %and the final error serves as the action.
  %%
  %Therefore, the optimal schedule follows from
  %the equation of motion of the particle.
  %
  We will show that
  the optimal schedule depends on the updating scheme
  (such as the single-bin scheme commonly used in WL
  or the Gaussian one popular in metadynamics),
  and the initial updating magnitude is ideally half of
  the previous equilibrium value.
  %
  Further,
  the single-bin updating scheme
  belongs to a class of bandpass updating schemes
  that are optimal for asymptotic convergence,
  and the optimal schedule of these schemes
  is the inverse number of sampling steps with a proper shift,
  as given by the $1/t$ prescription.
  %
\end{abstract}

@@

Many efficient flat-distribution-sampling (FDS) techniques
based on adaptive construction of the bias potential
have been introduced,
including the Wang-Landau (WL) algorithm\cite{
  wang2001, *wang2001pre}
and metadynamics\cite{huber1994,
*laio2002, *laio2008, *barducci2011, *sutto2012},
popular for MC and MD simulations, respectively\cite{junghans2014}.
%
%flat-distribution-sampling (FDS) techniques
%that build the bias potential adaptively or on the fly.
%
These techniques regularly update the bias potential
to discourage future visits to previously sampled configurations
by incrementally elevating the bias potential,
%and they are closely-related\cite{junghans2014, micheletti2004}.
%
%The bias potential is associated with discrete values of $z$.
%
A key difference lies
in the updating window function,
which specifies
a neighborhood around the current value of
$z$ on the free energy surface
where the updating should occur
as well as the relative updating magnitude.
%
Usually, the window function
takes the form of a discrete
$\delta$-function (a boxcar function one histogram bin wide)
in WL,
or that of a Gaussian
in metadynamics.
%
Although the two window functions
are not always associated with
the two methods\cite{micheletti2004, kim2006, *kim2007, junghans2014},
we will loosely use the terms WL and metadynamics
to represent FDS simulations using the
$\delta$- and Gaussian window functions, respectively.
%
%is that in the WL case,
%each updating step is limited to the bin
%containing the current $z$,
%while metadynamics adopts an extended
%Gaussian window
%covering several neighboring bins.
%%
%The latter is more often used in MD simulations
%as it guarantees a smooth profile
%of the bias potential
%suitable for numerical differentiation
%in deriving the bias force.
%
In both cases,
regular updates to the bias potential
%in both cases, the bias potential
%is updated frequently (usually every few MC or MD steps),
%and the fluctuation
disrupts the underlying equilibrium
sampling\cite{zhou2005, morozov2007, zhou2008},
and for convergence of the estimated PMF
one  has to gradually decrease
the updating magnitude of the bias.

The schedule of reducing
the updating magnitude,
therefore, critically affects
the precision of the final bias potential,
and hence that of the PMF\cite{
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
liang2007,
morozov2007, zhou2008,
komura2012, *caparica2012, *caparica2014,
barducci2008, dickson2011, dama2014}.
%
For the WL algorithm, the optimal updating schedule
is a function of the
inverse time\cite{
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
liang2007,
morozov2007, zhou2008}
or
the inverse number of simulation steps.
\note{
  About the origin of the inverse-time formula.
  Besides the Belardinelli paper,
  we can also find it in Eq. (6) of reference \cite{liang2007},
  if I am not too mistaken.
  But I find the paper is a bit hard to read.
  There, however, the formula is $t_0/t$ instead of $1/t$.
}
%
%The optimal schedule for metadynamics,
%however, is less studied.

In this study,
we present a method of computing
the optimal schedule
for an adaptive FDS method
with a general window function,
including the $\delta$- and Gaussian one
used in the WL algorithm and metadynamics.
%
Our method maps the optimization problem to a mechanical one,
in which the schedule plays the role of the velocity of
a free particle with a position-dependent mass,
and the final error becomes the action.
%
In this way, the optimal schedule
can be deduced from the equation of motion
of the particle that minimizes the action.
%
As the mass is defined from the updating window function,
the resulting optimal schedule
is different in the WL and metadynamics cases.
%
While the optimal schedule in the WL case
is given by the known inverse-time formula,
the optimal schedule for metadynamics is more complex
and sensitive to the simulation length
and the width of the Gaussian window.
%
In either case, however,
the initial updating magnitude
can be optimally set to
half of the previous equilibrium value,
which amounts to a shift of the origin of time in the WL case.
%
Further, we can generalize
the WL updating scheme
to a class of bandpass updating schemes
that are optimal for long simulations,
and the above origin-shifted inverse-time schedule
turns out to be optimal for these schemes.
%
%For finite-length simulations, however,
%metadynamics may deliver better performance.
%
These results may help to better
plan and optimize FDS simulations.

@@

%We will mainly focus on the adaptive methods below.
%
In the WL algorithm\cite{wang2001, *wang2001pre},
the bias potential, $V_i(t)$, is updated
at each sampling step $t$ as
%
\begin{equation}
  V_i(t+1)
  =
  V_i(t)
  +
  \delta_{i, \, i(t)}
  \frac{ \alpha(t) } { p_{i(t)} }
  ,
\label{eq:wl_update}
\end{equation}
%
where $i(t)$ is the bin at step $t$,
$\delta_{i, \, i(t)}$ is the Kronecker delta function,
which is $1$ when $i = i(t)$ or $0$ otherwise,
and $\alpha(t)$ is the updating magnitude.
%
We refer to this scheme as the \emph{single-bin scheme},
for it applies only to the bias potential
at the current bin, $i(t)$.
%
In a more general \emph{multiple-bin scheme}, however,
several neighboring bins are updated:
%
\begin{equation}
  V_i(t+1)
  =
  V_i(t)
  +
  w_{i, \, i(t)}
  \frac{ \alpha(t) }
       { p_{i(t)} },
  \label{eq:mbin_update}
\end{equation}
%
where the elements, $w_{ij}$'s, are a set of coefficients satisfying
the restrictions to be discussed
in Sec. \ref{sec:updating-matrix}.
%
The Gaussian updating scheme
used in metadynamics
is a multiple-bin scheme.

%One can show that with a constant magnitude,
%$\alpha(t) = \alpha > 0$,
%the distribution collected from
%an adaptive FDS simulation
%is identical to $p_i$ in the long time or asymptotic limit.
%
%However, as the above updates
%change $V_i(t)$ continuously,
%the equilibrium distribution,
%Eq. \eqref{eq:pi_p_V}, no longer holds
%at all times,
While adaptive FDS simulations
can often sample the desired distribution, $p_i$,
even with a constant magnitude,
$\alpha(t) = \alpha$,
frequent updates to the bias potential
disrupt the underlying equilibrium sampling
and introduce artificial errors that need to be
removed by gradually decreasing the updating magnitude\cite{
  belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
  zhou2005, morozov2007, zhou2008,
  laio2005, bussi2006, poulain2006, liang2007,
  crespo2010, *atchade2011, *fort2015}.
%%
%The source of the deviation is two-fold.
%%
%First, since $V_i(t)$ is updated continuously,
%there is a random noise that is proportional
%to $\sqrt \alpha$.
%%
%Second, there is a systematic error
%that comes from the updating dynamics itself,
%as it breaks the Markovian nature
%that underlies Eq. \eqref{eq:pi_p_v}.
%%

@@

In the original WL papers\cite{
wang2001, *wang2001pre},
the updating magnitude, $\alpha$,
(or $\ln f$ therein)
is controlled in stages.
%
In each stage, $\alpha$
is kept constant,
and the histogram along $i$
is collected and monitored.
%
Once the histogram is sufficiently flat,
we switch to a new stage
with $\alpha$ reduced by a factor of $1/2$\cite{
wang2001, *wang2001pre}.
%
While this scheme works well for early stages,
%
it tends to reduce $\alpha$
too quickly in late stages, making the asymptotic error
saturate\cite{
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016}.

%For the single-bin scheme
%used in WL algorithm,
A more effective way
of controlling the schedule, $\alpha(t)$,
in late stages
is to follow the formula
%
\begin{equation}
  \alpha(t) = \frac{1}{t},
  \label{eq:alpha_invt}
\end{equation}
%
where $t$,
referred to as the ``time'' below,
is the number of steps
from the beginning of the simulation\cite{
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
morozov2007, zhou2008,
komura2012, *caparica2012, *caparica2014}.
%
An interesting explanation of this result\cite{
  marsili2006, barducci2008}
comes from the interpretation of
the bias potential %under the WL updating scheme, Eq. \eqref{eq:wl_update},
as a runtime average computed from all data points collected so far,
so that the addition of a new data point only
affects the average by its due weight, $1/t$,
which is the inverse of the current sample size, $t$
(cf. Appendix \ref{sec:equilerr}).
%
Alternatively, we may derive it
from optimally planning the simulation lengths in the stages
and the reduction factor in the WL setup (cf. Appendix \ref{sec:wltoinvt}).

@@

%
For the histogram fluctuation part,
the average is zero by definition,
$\langle \zeta_i(t) \rangle = 0$,
%
and we will approximate it as white noise, i.e.
%The noise is not necessarily white,
%and the autocorrelation function
%depend only on the time difference
%
\begin{equation}
  \left\langle
    \zeta_i(t) \, \zeta_j(t')
  \right\rangle
  =
  %\sigma_{ij}(t - t')
  G^{(0)}_{ij} \, \delta(t - t')
  ,
  \notag
  %\label{eq:zeta_zeta_correlation}
\end{equation}
%
for some value of $G^{(0)}_{ij}$,
and $\delta(t)$ is the Dirac's delta function.
%where $\sigma_{ij}(t)$ is an even function
%that vanishes at large $t$.
%
This approximation is valid if the simulation period, $T$,
is much longer than the autocorrelation time of noise.
%
For non-white noise,
the equivalent value of $G^{(0)}_i$ can be approximately obtained
from the integral,
$\int_{-\infty}^\infty \sigma_{ij}(t) \, dt$,
or in the discrete time case,
$\sum_{-\infty}^\infty \sigma_{ij}(t)$.
%
\note{More explicitly,
if the transition matrix of a period, $\tau$,
is $T^\tau_{ij}$,
then
$
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  \langle h_i(t) \, h_j(t') \rangle
  -
  \langle h_i(t) \rangle \, \langle h_j(t') \rangle
  =
  T^{t - t'}_{ij} \, \pi_j - \pi_i \, \pi_j.
$
}%

@@

From Eqs.
\eqref{eq:v_def}-\eqref{eq:vbar_def},
\eqref{eq:rhox_sum}-\eqref{eq:h_ave},
%\eqref{eq:v_def},
%\eqref{eq:x_def},
%\eqref{eq:vbar_def},
%\eqref{eq:rhox_sum},
%\eqref{eq:vt_diffeq},
%\eqref{eq:h_split}
%\eqref{eq:h_ave},
%\eqref{eq:dvbardt},
we get a set of decoupled equations
%
\begin{equation}
  \dot x_i(t)
  =
  -\alpha(t) \, \left[ x_i(t) - \xi_i(t) \right]
  ,
\label{eq:dxdt_WL}
\end{equation}
%
where the shifted histogram fluctuation
%
\begin{equation}
  \xi_i(t)
  \equiv
  \frac{ \zeta_i(t) }
       { p_i }
  -
  \sum_{ j = 1 }^n
    \rho_j
    \frac{ \zeta_j(t) }
         {     p_j    }
  ,
\label{eq:xi_def}
\end{equation}
%
is also governed by white noise as is $\zeta_i(t)$,
i.e. we may write
\begin{equation}
  \langle \xi_i(t) \, \xi_i(t') \rangle
  = G_i \, \delta(t - t')
  ,
  \label{eq:G_def}
\end{equation}
for some $G_i$.
%
\note{The derivation:
$$
\begin{aligned}
  \dot x_i
  &\stackrel{\mathrm{Eq.\; \eqref{eq:x_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \dot v_i - \dot{ \bar v }
  \stackrel{\mathrm{Eqs.\; \eqref{eq:vt_diffeq}\; and\; \eqref{eq:vbar_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left(
    \frac{ h_i } { p_i }
    -
    \sum_{ j = 1 }^n
      \rho_j \frac{ h_j } { p_j }
  \right)
  \\
  &\stackrel{\mathrm{Eqs.\; \eqref{eq:h_split} \; and \; \eqref{eq:xi_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left(
    \frac{ \langle h_i \rangle } { p_i }
    -
    \sum_{ j = 1 }^n
      \rho_j \frac{ \langle h_j \rangle } { p_j }
    +
    \xi_i(t)
  \right)
  \\
  &\stackrel{\mathrm{Eqs.\; \eqref{eq:h_ave} \;and\; \eqref{eq:rhox_sum}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left[ -x_i(t) + \xi_i(t) \right].
\end{aligned}
$$
}

@@

We recall that
the formal solution of Eq. \eqref{eq:dxdt_WL} is
%
\begin{equation}
x_i(T)
=
x_i(0) \, e^{-q(T)}
+
\int_0^T
  \dot u\bigl( q(t) \bigr) \, \xi_i(t) \, dt,
\notag
%\label{eq:xt_solution}
\end{equation}
%
where
%
$
q(t) \equiv \int_0^t \alpha(t') \, dt',
$
%
and
%
\begin{align}
u(q')
&\equiv
e^{q' - q(T)}.
\label{eq:u_def}
\end{align}
%
At the end of period $T$,
the total error defined in Eq. \eqref{eq:error_def}
is given by
%
%\begin{align}
%  &\left\langle x_i^2(T) \right\rangle
%  =
%  \langle x_i^2(0) \rangle \, e^{-2 \, q(T)}
%  +
%  G_i \int_0^T
%    {\dot u}^2\bigl( q(t) \bigr) \,
%    \, dt.
%  \label{eq:x2t_average}
%\end{align}
%
\begin{align}
  E(T)
  =
  E(0) \, e^{-2 \, q(T)}
  +
  \Gamma \,
  \int_0^T
    {\dot u}^2\bigl( q(t) \bigr) \,
    \, dt
  ,
  \label{eq:ET_average}
\end{align}
where
\begin{equation}
  \Gamma \equiv \sum_{i=1}^n \rho_i \, G_i.
  \label{eq:Gamma_def}
\end{equation}

@@

Since the above optimization is performed
under a fixed $q(T)$,
it can be improved
by choosing a proper value of $q(T)$.
%
This is equivalent to choosing a value of $t_0$
\big[because $q(T) = \ln\bigl[1 + (T/t_0)\bigr]$,
from the $t = 0$ case of Eq. \eqref{eq:expqt}\big].
%
By using the optimal schedule in Eq. \eqref{eq:ET_average},
we find the final total error to be
%
$E(T) = [ E(0) \, t_0^2 + \Gamma \, T ]/(t_0 + T)^2$,
%
which is minimized if
%
\begin{equation}
t_0 = \frac{ \Gamma }
           { E(0) }
           ,
  \label{eq:t0_singlebin}
\end{equation}
%
and at this point, the error is given by
%
\begin{equation}
  E(T)
  =
  \frac{ \Gamma } { T + t_0 }
  .
  \label{eq:Emin_singlebin}
\end{equation}
%
This error is roughly equal to the overall histogram flatness
and thus it can be conveniently estimated\cite{zhou2005, zhou2008}
(see also Appendix \ref{sec:flatness}).

@@

The optimization procedure is similar to the single-bin case.
%
The key simplification comes from the
projection of the bias potential to the $n$ eigenmodes
of the updating matrix, $\mathbf w$,
formed by the relative magnitudes, $w_{ij}$'s.
%
%However, we will assume some restrictions and approximations
%to simplify the discussion.

@@

%It is worth pointing out
Finally, there is a stability condition.
%
In a \emph{stable} updating scheme,
no eigenmode grows indefinitely,
which means that no eigenvalue, $\lambda_k$,
can be negative.

@@

\subsubsection{Error functional}

>>>>

\begin{abstract}
  We present a method of computing the optimal schedule
  of the updating magnitude
  for a class of free energy calculation methods including
  the Wang-Landau (WL) algorithm and metadynamics.
  %
  These methods rely on adaptive construction of
  a bias potential that offsets
  the potential of mean force by histogram-based adaptive updates.
  %
  The updating magnitude should decrease over time
  to reduce the noise in the bias potential,
  and one may speed up the convergence by choosing an optimal schedule
  of the updating magnitude.
  %
  %Here, we will show that the optimal schedule
  %can be derived
  %from a mechanical analogy, in which
  %the schedule corresponds to
  %the velocity of a free particle with a position-dependent mass
  %and the final error serves as the action.
  %%
  %Therefore, the optimal schedule follows from
  %the equation of motion of the particle.
  %
  We will show that
  the optimal schedule depends on the updating scheme
  (such as the single-bin scheme commonly used in WL
  or the Gaussian one popular in metadynamics),
  and the initial updating magnitude is ideally half of
  the previous equilibrium value.
  %
  Further,
  the single-bin updating scheme
  belongs to a class of bandpass updating schemes
  that are optimal for asymptotic convergence,
  and the optimal schedule of these schemes
  is the inverse number of updating steps with a proper shift,
  as given by the $1/t$ prescription.
  %
\end{abstract}

@@

Many efficient flat-distribution-sampling (FDS) techniques
based on adaptive construction of the bias potential
have been introduced,
including the Wang-Landau (WL) algorithm\cite{
  wang2001, *wang2001pre}
and metadynamics\cite{huber1994,
*laio2002, *laio2008, *barducci2011, *sutto2012},
popular for MC and MD simulations, respectively\cite{junghans2014}.
%
%flat-distribution-sampling (FDS) techniques
%that build the bias potential adaptively or on the fly.
%
These techniques regularly update the bias potential
to discourage future visits to previously sampled configurations
by incrementally elevating the bias potential,
%and they are closely-related\cite{junghans2014, micheletti2004}.
%
%The bias potential is associated with discrete values of $z$.
%
A key difference lies
in the updating window function,
which specifies
a neighborhood around the current value of
$z$ on the free energy surface
where the updating should occur
as well as the relative updating magnitude.
%
Usually, the window function
takes the form of a discrete
$\delta$-function (a boxcar function one histogram bin wide)
in WL,
or that of a Gaussian
in metadynamics.
%
Although the two window functions
are not always associated with
the two methods\cite{micheletti2004, kim2006, *kim2007, junghans2014},
we will colloquially use the terms WL and metadynamics
to represent FDS simulations using the
single-bin and Gaussian window functions, respectively.
%
%is that in the WL case,
%each updating step is limited to the bin
%containing the current $z$,
%while metadynamics adopts an extended
%Gaussian window
%covering several neighboring bins.
%%
%The latter is more often used in MD simulations
%as it guarantees a smooth profile
%of the bias potential
%suitable for numerical differentiation
%in deriving the bias force.
%
In both cases,
regular updates to the bias potential
%in both cases, the bias potential
%is updated frequently (usually every few MC or MD steps),
%and the fluctuation
disrupts the underlying equilibrium
sampling\cite{zhou2005, morozov2007, zhou2008},
and for convergence of the estimated PMF
one  has to gradually decrease
the updating magnitude of the bias.

The schedule of reducing
the updating magnitude,
therefore, critically affects
the precision of the final bias potential,
and hence that of the PMF\cite{
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
liang2007,
morozov2007, zhou2008,
komura2012, *caparica2012, *caparica2014,
barducci2008, dickson2011, dama2014}.
%
For the single-bin (WL) case, the optimal updating schedule
is a function of the
inverse time\cite{
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
liang2007,
morozov2007, zhou2008}
or
the inverse number of updating steps.
\note{
  About the origin of the inverse-time formula.
  Besides the Belardinelli paper,
  we can also find it in Eq. (6) of reference \cite{liang2007},
  if I am not too mistaken.
  But I find the paper is a bit hard to read.
  There, however, the formula is $t_0/t$ instead of $1/t$.
}
%
%The optimal schedule for metadynamics,
%however, is less studied.

In this study,
we present a method of computing
the optimal schedule
for an adaptive FDS method
with a general window function,
including the single-bin and Gaussian one
used in the WL algorithm and metadynamics.
%
Our method maps the optimization problem to a mechanical one,
in which the schedule plays the role of the velocity of
a free particle with a position-dependent mass,
and the final error becomes the action.
%
In this way, the optimal schedule
can be deduced from the equation of motion
of the particle that minimizes the action.
%
The resulting optimal schedule
depends on the updating window function
through the mass function.
%
While the optimal schedule in the single-bin case
is given by the known inverse-time formula,
that for Gaussian case is more complex
and sensitive to the simulation length
and the width of the Gaussian window.
%
In either case, however,
the initial updating magnitude
can be optimally set to
half of the previous equilibrium value,
which amounts to a shift of the origin of time
in the single-bin case.
%
Further, we can generalize
the single-bin updating scheme
to a class of bandpass updating schemes
that are optimal for long simulations,
and the above origin-shifted inverse-time schedule
turns out to be optimal for these schemes.
%
%For finite-length simulations, however,
%metadynamics may deliver better performance.
%
These results may help to better
plan and optimize FDS simulations.

@@

%We will mainly focus on the adaptive methods below.
%
In the WL algorithm\cite{wang2001, *wang2001pre},
the bias potential, $V_i(t)$, is updated
at each sampling step $t$ as
%
\begin{equation}
  V_i(t+1)
  =
  V_i(t)
  +
  \delta_{i, \, i(t)}
  \frac{ \alpha(t) } { p_{i(t)} }
  ,
\label{eq:wl_update}
\end{equation}
%
where $i(t)$ is the bin at step $t$,
$\delta_{i, \, i(t)}$ is the Kronecker delta function,
which is $1$ when $i = i(t)$ or $0$ otherwise,
and $\alpha(t)$ is the updating magnitude.
%
We refer to this scheme as the \emph{single-bin scheme},
for it applies only to the bias potential
at the current bin, $i(t)$.
%
In a more general \emph{multiple-bin scheme},
including the Gaussian updating scheme
used in metadynamics,
several neighboring bins are updated:
%
\begin{equation}
  V_i(t+1)
  =
  V_i(t)
  +
  w_{i, \, i(t)}
  \frac{ \alpha(t) }
       { p_{i(t)} },
  \label{eq:mbin_update}
\end{equation}
%
where the elements, $w_{ij}$'s, are a set of coefficients satisfying
the restrictions to be discussed
in Sec. \ref{sec:updating-matrix}.
%
Note that if the updating occurs
only every few sampling steps,
then $t$ means the number of updating steps.

@@

While adaptive FDS simulations
can often sample the desired distribution,
frequent updates to the bias potential
disrupt the underlying equilibrium sampling
and introduce artificial errors that need to be
removed by gradually decreasing the updating magnitude\cite{
  belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
  zhou2005, morozov2007, zhou2008,
  laio2005, bussi2006, poulain2006, liang2007,
  crespo2010, *atchade2011, *fort2015}.
%One can show that with a constant magnitude,
%$\alpha(t) = \alpha > 0$,
%the distribution collected from
%an adaptive FDS simulation
%is identical to $p_i$ in the long time or asymptotic limit.
%
%However, as the above updates
%change $V_i(t)$ continuously,
%the equilibrium distribution,
%Eq. \eqref{eq:pi_p_V}, no longer holds
%at all times,
%%
%The source of the deviation is two-fold.
%%
%First, since $V_i(t)$ is updated continuously,
%there is a random noise that is proportional
%to $\sqrt \alpha$.
%%
%Second, there is a systematic error
%that comes from the updating dynamics itself,
%as it breaks the Markovian nature
%that underlies Eq. \eqref{eq:pi_p_v}.
%%

In the original WL papers\cite{
wang2001, *wang2001pre},
the updating magnitude, $\alpha(t)$,
(or $\ln f$ therein)
is controlled in stages.
%
In each stage, $\alpha(t)$
is kept constant,
and the histogram along $i$
is collected and monitored.
%
Once the histogram is sufficiently flat,
we switch to a new stage
with $\alpha(t)$ reduced by a factor of $1/2$\cite{
wang2001, *wang2001pre}.
%
While this scheme works well for early stages,
it tends to reduce the magnitude
too quickly in late stages, making the asymptotic error
saturate\cite{
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016}.

%For the single-bin scheme
%used in WL algorithm,
A more effective way
of controlling the schedule, $\alpha(t)$,
in late stages
is to follow the formula
%
\begin{equation}
  \alpha(t) = \frac{1}{t},
  \label{eq:alpha_invt}
\end{equation}
%
where $t$,
referred to as the ``time'' below,
is the number of updating steps
from the beginning of the simulation\cite{
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
morozov2007, zhou2008,
komura2012, *caparica2012, *caparica2014}.
%
An intuitive explanation of this result\cite{
  marsili2006, barducci2008}
comes from the interpretation of
the bias potential %under the WL updating scheme, Eq. \eqref{eq:wl_update},
as a runtime average computed from all data points collected so far,
so that the addition of a new data point only
affects the average by its due weight, $1/t$,
which is the inverse of the current sample size, $t$
(cf. Appendix \ref{sec:equilerr}).
%
Alternatively, we may derive it
from optimally planning the simulation lengths in the stages
and the reduction factor in the WL setup (cf. Appendix \ref{sec:wltoinvt}).

@@

%
For the histogram fluctuation part,
the average is zero by definition,
$\langle \zeta_i(t) \rangle = 0$,
%
and we will approximate it as white noise, i.e.
%The noise is not necessarily white,
%and the autocorrelation function
%depend only on the time difference
%
\begin{equation}
  \left\langle
    \zeta_i(t) \, \zeta_j(t')
  \right\rangle
  =
  %\sigma_{ij}(t - t')
  G^{(0)}_{ij} \, \delta(t - t')
  ,
  \notag
  %\label{eq:zeta_zeta_correlation}
\end{equation}
%
for some value of $G^{(0)}_{ij}$,
and $\delta(t)$ is the Dirac's delta function.
%where $\sigma_{ij}(t)$ is an even function
%that vanishes at large $t$.
%
This approximation is valid if the simulation period, $T$,
is much longer than the autocorrelation time of noise.
%
For non-white noise,
the equivalent value of $G^{(0)}_i$ can be approximately obtained
%from the integral,
%$\int_{-\infty}^\infty \sigma_{ij}(t) \, dt$,
%or in the discrete time case,
$\sum_{t = -\infty}^\infty \sigma_{ij}(t)$.
%
\note{More explicitly,
if the transition matrix of a period, $\tau$,
is $T^\tau_{ij}$,
then
$
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  \langle h_i(t) \, h_j(t') \rangle
  -
  \langle h_i(t) \rangle \, \langle h_j(t') \rangle
  =
  T^{t - t'}_{ij} \, \pi_j - \pi_i \, \pi_j.
$
}%

@@

From Eqs.
\eqref{eq:v_def}-\eqref{eq:vbar_def},
\eqref{eq:rhox_sum}-\eqref{eq:h_ave},
%\eqref{eq:v_def},
%\eqref{eq:x_def},
%\eqref{eq:vbar_def},
%\eqref{eq:rhox_sum},
%\eqref{eq:vt_diffeq},
%\eqref{eq:h_split}
%\eqref{eq:h_ave},
%\eqref{eq:dvbardt},
we get % a set of decoupled equations
%
\begin{equation}
  \dot x_i(t)
  =
  -\alpha(t) \, \left[ x_i(t) - \xi_i(t) \right]
  ,
\label{eq:dxdt_singlebin}
\end{equation}
%
where the shifted histogram fluctuation
%
\begin{equation}
  \xi_i(t)
  \equiv
  \frac{ \zeta_i(t) }
       { p_i }
  -
  \sum_{ j = 1 }^n
    \rho_j
    \frac{ \zeta_j(t) }
         {     p_j    }
  ,
\label{eq:xi_def}
\end{equation}
%
is also governed by white noise as is $\zeta_i(t)$:
%i.e. we may write
\begin{equation}
  \langle \xi_i(t) \, \xi_i(t') \rangle
  = G_i \, \delta(t - t')
  ,
  \label{eq:G_def}
\end{equation}
for some $G_i$.
%
\note{The derivation:
$$
\begin{aligned}
  \dot x_i
  &\stackrel{\mathrm{Eq.\; \eqref{eq:x_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \dot v_i - \dot{ \bar v }
  \stackrel{\mathrm{Eqs.\; \eqref{eq:vt_diffeq}\; and\; \eqref{eq:vbar_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left(
    \frac{ h_i } { p_i }
    -
    \sum_{ j = 1 }^n
      \rho_j \frac{ h_j } { p_j }
  \right)
  \\
  &\stackrel{\mathrm{Eqs.\; \eqref{eq:h_split} \; and \; \eqref{eq:xi_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left(
    \frac{ \langle h_i \rangle } { p_i }
    -
    \sum_{ j = 1 }^n
      \rho_j \frac{ \langle h_j \rangle } { p_j }
    +
    \xi_i(t)
  \right)
  \\
  &\stackrel{\mathrm{Eqs.\; \eqref{eq:h_ave} \;and\; \eqref{eq:rhox_sum}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left[ -x_i(t) + \xi_i(t) \right].
\end{aligned}
$$
}

@@

We recall that
the formal solution of Eq. \eqref{eq:dxdt_singlebin} is
%
\begin{equation}
x_i(T)
=
x_i(0) \, e^{-q(T)}
+
\int_0^T
  \dot u\bigl( q(t) \bigr) \, \xi_i(t) \, dt,
\notag
%\label{eq:xt_solution}
\end{equation}
%
where
%
$
q(t) \equiv \int_0^t \alpha(t') \, dt',
$
%
and
%
\begin{align}
u(q')
&\equiv
e^{q' - q(T)}.
\label{eq:u_def}
\end{align}
%
At the end of period $T$,
the total error defined in Eq. \eqref{eq:error_def}
is given by
%
%\begin{align}
%  &\left\langle x_i^2(T) \right\rangle
%  =
%  \langle x_i^2(0) \rangle \, e^{-2 \, q(T)}
%  +
%  G_i \int_0^T
%    {\dot u}^2\bigl( q(t) \bigr) \,
%    \, dt.
%  \label{eq:x2t_average}
%\end{align}
%
\begin{align}
  E(T)
  =
  E(0) \, e^{-2 \, q(T)}
  +
  \Gamma \,
  \int_0^T
    {\dot u}^2\bigl( q(t) \bigr) \,
    \, dt
  ,
  \label{eq:ET_average}
\end{align}
where
\begin{equation}
  \Gamma \equiv \sum_{i=1}^n \rho_i \, G_i.
  \label{eq:Gamma_def}
\end{equation}

@@

We now determine the optimal value of $q(T)$,
which is fixed in the above process.
%
This is equivalent to choosing a value of $t_0$
\big[because $q(T) = \ln\bigl[1 + (T/t_0)\bigr]$,
from the $t = 0$ case of Eq. \eqref{eq:expqt}\big].
%
By using Eq. \eqref{eq:alpha_invt} in Eq. \eqref{eq:ET_average},
we find the final total error to be
%
$E(T) = [ E(0) \, t_0^2 + \Gamma \, T ]/(t_0 + T)^2$,
%
which is minimized if
%
\begin{equation}
t_0 = \frac{ \Gamma }
           { E(0) }
           ,
  \label{eq:t0_singlebin}
\end{equation}
%
and at this point, the error is given by
%
\begin{equation}
  E(T)
  =
  \frac{ \Gamma } { T + t_0 }
  .
  \label{eq:Emin_singlebin}
\end{equation}
%
This error is roughly equal to
the overall histogram flatness\cite{zhou2005, zhou2008}
and thus it can be conveniently estimated
(see also Appendix \ref{sec:flatness}).

@@

The optimization procedure is similar to the single-bin case.
%
The key step comes from the
projection of the bias potential to the $n$ eigenmodes
of the updating matrix, $\mathbf w$,
formed by the relative magnitudes, $w_{ij}$'s.
%
%However, we will assume some restrictions and approximations
%to simplify the discussion.

@@

%It is worth pointing out
Finally, there is a stability condition.
%
In a \emph{stable} updating scheme,
no eigenmode grows indefinitely,
which means that, as we will show, no eigenvalue, $\lambda_k$,
can be negative.

@@

\subsubsection{Error}

The error defined in Eq. \eqref{eq:error_def}
can be rewritten in terms of the eigenmodes of $\mathbf w$,
%
\begin{align}
  \Err
  \equiv
  \sum_{i = 1}^n \rho_i \,
    \left\langle x_i^2 \right\rangle
  =
  \sum_{i, j, k=1}^n \phi_{ki} \, \phi_{kj} \,
    \left\langle x_i \, x_j \right\rangle
  =
  \sum_{k = 1}^n
    \left\langle
      y_k^2
    \right\rangle
  ,
\notag
\end{align}
%
where we have used Eq. \eqref{eq:eig_orthonormal_cols}.
%
From Eq. \eqref{eq:eigenmode1},
the first eigenmode,
$y_1 = \sum_{j=1}^n \rho_j \, x_j$,
vanishes identically by Eq. \eqref{eq:rhox_sum}.
%
Thus, we can start the last sum %in Eq. \eqref{eq:y2_sum}
from $k = 2$,
%
\begin{equation}
  \Err(T)
  =
  \sum_{k = 2}^n
    \left\langle
    y_k^2(T)
    \right\rangle
  .
\label{eq:y2_sum}
\end{equation}
